MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-07-13 latex/paper/sections/11-placement_results.tex
start future work

\vspace{0.5cm}
{
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/results/utilization.png}
    \captionof{figure}{Utilization of BELs on a \texttt{xc7z020} FPGA for a FIR filter design}
    \label{fig:utilization}
}
\vspace{0.5cm}

@@

{
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/results/combined_cost_history_linear.png}
    \captionof{figure}{All placers overlaid}
    \label{fig:placers_overlay}
}

@@

{
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/placement/cooling_schedule_10000_98.png}
    \captionof{figure}{Cooling Schedule: \(T_0=10000\), \(\alpha=0.98\)}
    \label{fig:cooling_schedule_10000_98}
}

\end{multicols}
{
    \centering
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/random_placement.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000010.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000100.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000299.png}
    \captionof{figure}{\texttt{PlacerGreedyRandom}}
    \label{fig:PGRSnapshots}
}

@@

    Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.

    The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock trees. 
    Since clock nets tend to have very high fanout, they can use up a significant portion of the routing resource if they are routed through general routing.
    Modern FPGAs like the 7-Series have dedicated clock routes which can efficiently distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
    These special clock routes are accessed by routing the off-chip crystal oscillator clock to one of these clock buffers which connects to the clock tree network.
    It is always best practice to use them to improve the routability of the design, and simultaneously, to reduce clock skew and improve timing. 
    Our simple FIR Filter design only operates on a single clock domain, so only one clock buffer \texttt{Site} is used. 
    An HDL design operating on many clock domains will use more clock buffers \texttt{Site}s or a dedicated clock generator such as a phase locked loop (PLL) \texttt{Site} or mixed-mode clock manager (MCMM) \texttt{Site}.

    The right and left edges of the FPGA device is where most of the general purpose IO (GPIO) is located. 
    In the top right of the device is where we source our reset signal. 
    Like the clock net, the reset net also has high fanout and is supplied to all synchronous resources.
    Unlike clock signals, reset signals are not subject to sensitive timing constraints as they are not active on most clock cycles and can tolerate higher skew.
    Thus, they are simply routed via general routing.

    Observe how \texttt{PlacerGreedyMidpoint} \ref{fig:PGMSnapshots} and \texttt{PlacerAnnealMidpoint} \ref{fig:PAMSnapshots} both show almost not shrinkage in profile after the first 10 iterations, likely owing to the inability to hill-climb.

>>>>

\vspace{0.25cm}
{
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/results/utilization.png}
    \captionof{figure}{Utilization of BELs on a \texttt{xc7z020} FPGA for a FIR filter design}
    \label{fig:utilization}
}
\vspace{0.25cm}

@@

{
    \centering
    \includegraphics[width=\columnwidth]{figures/results/combined_cost_history_linear.png}
    \captionof{figure}{All placers overlaid}
    \label{fig:placers_overlay}
}

@@

{
    \centering
    \includegraphics[width=0.6\columnwidth]{figures/placement/cooling_schedule_10000_98.png}
    \captionof{figure}{Cooling Schedule: \(T_0=10000\), \(\alpha=0.98\)}
    \label{fig:cooling_schedule_10000_98}
}

\newpage
Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.

The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock trees. 
Since clock nets tend to have very high fanout, they can use up a significant portion of the routing resources if they are routed through general routing.
Modern FPGAs like the 7-Series have dedicated clock routes which can efficiently distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
These special clock routes are accessed by routing the raw clock signal from the crystal oscillator IO port to one these on-chip clock buffers which connects to the on-chip clock tree network.
It is always best practice to use them to improve the routability of the design, and simultaneously, to reduce clock skew and improve timing. 
Our simple FIR Filter design only operates on a single clock domain, so only one clock buffer \texttt{Site} is used. 
An HDL design operating on many clock domains will use more clock buffers or a dedicated clock generator such as a phase locked loop (PLL) or mixed-mode clock manager (MCMM).

The right and left edges of the FPGA device is where most of the general purpose I/O (GPIO) is located. 
In the top right of the device is where we source our reset signal. 
Like the clock net, the reset net also has high fanout and is supplied to all synchronous elements.
Unlike clock signals, reset signals are not subject to sensitive timing constraints as they are not active on most clock cycles and can tolerate higher skew.
Thus, they are simply routed via general routing.

Observe how \texttt{PlacerGreedyMidpoint} \ref{fig:PGMSnapshots} and \texttt{PlacerAnnealMidpoint} \ref{fig:PAMSnapshots} both show almost not shrinkage in profile after the first 10 iterations, likely owing to the inability to hill-climb, while the other three placers show incremental shrinkage.

Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the large grid of \texttt{SLICEL}s (cyan).

\newpage
\end{multicols}
{
    \centering
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/random_placement.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000010.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000100.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000299.png}
    \captionof{figure}{\texttt{PlacerGreedyRandom}}
    \label{fig:PGRSnapshots}
}

@@

### 2025-07-13 latex/paper/sections/12-future_work.tex
start future work

\section{Future Work}
In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} homogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area.

There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow.

The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.

>>>>

\section{Future Work}

\subsection{Expand Library of Primitives}
In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} homogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area.

\subsection{Robustness}
In its current state, the prepacker and packer struggle with data busses larger than 24-bits, especially with DSP functions like addition and multiplication.

\subsection{Improvements to Packing}
There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow. 
The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.

\subsection{Improvements to SA}

\subsection{Force-Directed and Analytical Placement}

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
We will implement variations on the vanilla SA algorithm that use centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.

Directed moves as opposed to undirected moves give rise to the concept of legalization. 
When using undirected moves as in SA, we can simply keep track of a \texttt{List} or \texttt{HashMap} of occupied \texttt{Site}s and available \texttt{Site}s and select from that data structure via random selection. 
However, for directed moves like centroid selection, the centroid between \texttt{Site}s may not necessarily fall directly on another discrete \texttt{Site} on the \texttt{device}. 
Even when it does, the centroid \texttt{Site} may be incompatible with the \texttt{SiteInst} or may fail to meet some other placement constraint. 
In such cases, we must do a neighborhood search around the centroid up to a certain radius to find a \texttt{Site} that satisfies all the constraints of the \texttt{SiteInst} at hand. 
We will take some inspiration from ray-marching in graphics and do a diamond-shaped spiral search around the centroid where we check the legality of each \texttt{Site} as we step through the spiral and select the first \texttt{Site} that meets all of the placement constraints.

In force directed placers, there is a concept of global placement followed by detailed placement.
Global placement places object instances freely on an analog plane and is driven purely by directed movement and analytical solution.
In global placement stage, only non-hardware related objectives are considered. 
Object instance placements do not have to adhere to any hardware constraints and can even overlap with one another.
Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the instances onto the discrete grid of resources of the \texttt{device}.
Overlapping instances are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy. 
There also exists other strategies to mix packing, placement, and legalization in different orders, such as those studied in \cite{ExplicitPacking}. 
In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.
